---
title: "FINAL PROJECT (BMIN503/EPID60) /// Preliminary_Results"
author: "David Karp (email: dkarp@upenn.edu; github: [dnkarp](https://github.com/dnkarp))"
date: "12/06/2017"
output:
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---

```{r setup, echo=FALSE, cache=FALSE}
options(width = 400)
require("knitr")
opts_chunk$set(fig.width=7, fig.height=4.1) 
opts_knit$set(root.dir = "~/Dropbox/courses/fall-2017/EPID-600/final_project/BMIN503_Final_Project")

```  

***

- IN-CLASS PRESENTATION: 11/30/17
- FINAL REPORT: 12/08/17

> GEOGRAPHY OF THE US OPIOID CRISIS: SPACE-TIME ANALYSIS USING R

##PART 1: DESCRIPTION

+[1a]Set-up Environment
```{r set environment, eval=FALSE}  
###--ENVIRONMENT--###----

#RData files for saving/restoring working environments 
#save.image("prelim_results--120617.RData")
#load("prelim_results--120617.RData")

```

+[1b]Load Libraries
```{r load libraries, eval=FALSE echo=FALSE} 
###--LIBRARIES--###----

require("rgdal")
require("maptools")
require("ggplot2")
require("plyr")
require("dplyr")
require("tigris")
require("data.table")
require("tmap")
require("tableone")
require("spdep")
require("maptools")
require("classInt")
require("gstat")
require("randomForest")
require("gplots")
require("RColorBrewer")

```

+[1c]Pre-process Data
```{r pre-process data, eval=FALSE}
###--LOAD/CLEAN OUTCOMES (MCOD) DATA--###----

#source("data/multiple_cause_of_death/mcod.R")

```

+[1d]Load Data
```{r load data, eval=FALSE}
###--LOAD OUTCOMES (MCOD) DATA--###----

#load cleaned data, incident level multiple casue of death data
#mcod.allyears <- readRDS(file="data/multiple_cause_of_death/mcod_95-15.RDa")

#NOTE: Incident level data cannot be pushed to Github's servers, various portion of data are commented out where these data are used

```

+[1e]Generate Descriptive Statistics
```{r descriptive stats, eval=TRUE}
###--TIME TRENDS: LINE PLOTS--###----

#copy as working data set
#dt.allyears <- mcod.allyears

#review variable names
#names(dt.allyears)

#group by year and demographics
#dat.age <- dt.allyears %>% group_by(year, age.bins) %>%
#  summarise_at(.vars = c("drug.all", "drug.rx","drug.opioid","drug.heroin"), funs(sum))
#saveRDS(dat.age,file="data/multiple_cause_of_death/dat_age.RDa")
dat.age <- readRDS(file="data/multiple_cause_of_death/dat_age.RDa")

#dat.sex <- dt.allyears %>% group_by(year, sex) %>%
#  summarise_at(.vars = c("drug.all", "drug.rx","drug.opioid","drug.heroin"), funs(sum))
#saveRDS(dat.sex,file="data/multiple_cause_of_death/dat_sex.RDa")
dat.sex <- readRDS(file="data/multiple_cause_of_death/dat_sex.RDa")

#dat.race <- dt.allyears %>% group_by(year, race) %>%
#  summarise_at(.vars = c("drug.all", "drug.rx","drug.opioid","drug.heroin"), funs(sum))
#saveRDS(dat.race,file="data/multiple_cause_of_death/dat_race.RDa")
dat.race <- readRDS(file="data/multiple_cause_of_death/dat_race.RDa")

#dat.heroin <- dt.allyears[drug.opioid==1 | drug.heroin==1] %>% group_by(year, drug.heroin) %>%
#  summarise_at(.vars = c("drug.all"), funs(sum)) #generate binary of heroin v. opioid (heroin=1, opioid=0)
#saveRDS(dat.heroin,file="data/multiple_cause_of_death/dat_heroin.RDa")
dat.heroin <- readRDS(file="data/multiple_cause_of_death/dat_heroin.RDa")

#generate exhibits 1-4 // OPTIONAL print to file directory ("./results")
ggplot(dat.age, aes(x=year, y=drug.all, group=age.bins)) +
  geom_line(aes(color=age.bins))+
  geom_point(aes(color=age.bins))
#dev.print(png, file = "results/EX1-drug-all-year-by-age.png", width = 500, height = 300) #print to file

ggplot(dat.sex, aes(x=year, y=drug.all, group=sex)) +
  geom_line(aes(color=sex))+
  geom_point(aes(color=sex))
#dev.print(png, file = "results/EX2-drug-all-year-by-sex.png", width = 500, height = 300) #print to file

ggplot(dat.race, aes(x=year, y=drug.all, group=race)) +
  geom_line(aes(color=race))+
  geom_point(aes(color=race))
#dev.print(png, file = "results/EX3-drug-all-year-by-race.png", width = 500, height = 300) #print to file

ggplot(dat.heroin, aes(x=year, y=drug.all, group=drug.heroin)) +
  geom_line(aes(color=drug.heroin))+
  geom_point(aes(color=drug.heroin))
#dev.print(png, file = "results/EX4-drug-all-year-by-heroin.png", width = 500, height = 300) #print to file


```

+[1e]Generate Descriptive Statistics
```{r descriptive stats, eval=TRUE}
###--SUMMARY OF POPULATION: TABLE ONE--###----

#dat <- dt.allyears[(dt.allyears$drug.heroin == 1 | dt.allyears$drug.opioid == 1),] %>% 
#  mutate(heroin = ifelse(!(drug.heroin == 1 | drug.opioid == 1), NA, drug.heroin))

#tableOne.year <- CreateTableOne(strata = "year", vars = c("heroin","sex","race","hispanic","age","age.bins","placeofdeath"), data = dat, factorVars = c("heroin"))

#tableOne.heroin <- CreateTableOne(strata = "heroin", vars = c("year","sex","race","hispanic","age.bins","placeofdeath"), data = dat)


#tableOne.year #print table
#tableOne.heroin #print table


```

+[1f]Generate Age-adjusted Rates by Geography
```{r pre-process data, eval=FALSE}
###--GENERATE AGE-ADJSUTED OUTCOMES (RATES BY GEOGRAPHY)--###----

#source("data/censusapi/pull_census_vars_120417.R")
#source("data/multiple_cause_of_death/age_adjust_rates.R")

```

+[1g]Load Age-adjusted Data
```{r load data, eval=FALSE}
###--IMPORT AGE-ADJUSTED OUTCOMES DATA--###----

#county
#mcod.cnty <- readRDS(file="../local_data/multiple_cause_of_death/aa.mcod.cnty.allyrs.RDa") #load unsuppressed, all years
mcod.cnty <- readRDS(file="data/multiple_cause_of_death/aa_mcod_cnty_2015_suppressed.RDa")
mcod.cnty <- mcod.cnty %>% mutate(stateFIPS=substr(geo,1,2)) %>% mutate(countyFIPS=geo)
                                                                       
#state
#mcod.st <- readRDS(file="../local_data/multiple_cause_of_death/aa.mcod.st.allyrs.RDa") #load unsuppressed, all years
mcod.st <- readRDS(file="data/multiple_cause_of_death/aa_mcod_st_2015_suppressed.RDa")
mcod.st <- mcod.st %>% mutate(stateFIPS=geo)

#national
#mcod.ntl <- readRDS(file="data/multiple_cause_of_death/aa.mcod.ntl.allyrs.RDa")
#mcod.ntl <- mcod.ntl %>% mutate(geo=1)

mkwide <- function(x){
  d<-setDT(x[,1:10])
  vars <- names(d[3:10])
  return(dcast(d, geo ~ substr(year,3,4), value.var = vars) %>% select(-starts_with("geo.1"),-starts_with("year.1")))

}

#make wide formats
mcod.cnty.wide <- mkwide(mcod.cnty) %>% mutate(stateFIPS=substr(geo,1,2)) %>% mutate(countyFIPS=geo) %>% mutate(GEOID=geo)
mcod.st.wide <- mkwide(mcod.st) %>% mutate(stateFIPS=geo) %>% mutate(GEOID=geo)
#mcod.ntl.wide <- mkwide(mcod.ntl)

#percent change (for future iterations with multiple year format)
#mcod.st.wide[, all.pch9515 := (100*round((all_15-all_95)/all_95,3))]
#mcod.st.wide[, rx.pch9515 := (100*round((rx_15-rx_95)/rx_95,3))]
#mcod.st.wide[, opioid.pch9515 := (100*round((opioid_15-opioid_95)/opioid_95,3))]
#mcod.st.wide[, heroin.pch9515 := (100*round((heroin_15-heroin_95)/heroin_95,3))]

#mcod.cnty.wide[, all.pch9515 := (100*round((all_15-all_95)/all_95,3))]
#mcod.cnty.wide[, rx.pch9515 := (100*round((rx_15-rx_95)/rx_95,3))]
#mcod.cnty.wide[, opioid.pch9515 := (100*round((opioid_15-opioid_95)/opioid_95,3))]
#mcod.cnty.wide[, heroin.pch9515 := (100*round((heroin_15-heroin_95)/heroin_95,3))]


```

+[1h]Load Geographic Data
```{r load data, eval=FALSE}
###--LOAD GEOGRAPHIC DATA--###----

#source("data/geography/load_geographies.R") #load source geographies from tigris

#load county geographies
geo.cnty <- readRDS(file="data/geography/UScountySPDF.RDa") #loads all counties

#load state geographies
geo.st <- readRDS(file="data/geography/USstateSPDF.RDa") #load all states

#specifiy codes for outlying territories & lower 48
territories <- c("60","64","66","68","69","70","72","74","78")
AK_and_HI <- c("02","15")

#create place name
USPSlist<-geo.st@data[,c("STUSPS","GEOID")]
geo.cnty<-merge(geo.cnty,USPSlist,by.x="STATEFP",by.y="GEOID",x.all=T)
geo.cnty@data <- geo.cnty@data %>% mutate(NAME=paste(NAMELSAD,STUSPS,sep=", "))

```

+[1i]Merge Data
```{r merge data, eval=TRUE}
###--MERGE GEO+OUTCOMES DATA--###----

#specify subsets by year
mcod.cnty15 <- mcod.cnty[mcod.cnty$year=="2015",] %>% mutate(GEOID=countyFIPS) #subset county outcomes 2015 only
mcod.st15 <- mcod.st[mcod.st$year=="2015",] %>% mutate(GEOID=stateFIPS) #subset state outcomes 2015 only

#to specify geogrpahic subsets
#geo.st.sub <- geo.st[geo.st$STUSPS %in% c("PA","NY","NJ","MD","WV","OH","DE","VA","CT"),] #specify a subset

## (for SpatialDataFrames) use geo_join
#state join
geo.st <- geo.st[!(geo.st$STATEFP %in% c(AK_and_HI, territories)),] #exclude territories,AK,HI
geo.st.mcod <- geo_join(geo.st, mcod.st.wide, by="GEOID", how="left") #join outcomes data to state geographies

#county join
geo.cnty <- geo.cnty[!(geo.cnty$STATEFP %in% c(AK_and_HI, territories)),] #exclude territories,AK,HI 
geo.cnty.mcod <- geo_join(geo.cnty, mcod.cnty.wide, by="GEOID", how="left") #join outcomes data to county geographies

#names(geo.cnty.mcod)
#geo.cnty.mcod@data[is.na(geo.cnty.mcod$all_15),]$all_15 <- 0
#names(geo.st.mcod)
#geo.st.mcod@data[is.na(geo.st.mcod$all_15),]$all_15 <- 0

```

+[1j]Get summary statistics
```{r descriptive stats, eval=TRUE}
###--GEOGRAPHIC VARIATION: HISTOGRAMS + BOX PLOTS--###----

#//TODO//#
#stratify by drug
#compare variation by year

#get summary distribution
summary(mcod.st15$all)

#plot histograms (state level outcomes)
ggplot(data=mcod.st15, aes(x=all)) +
    geom_histogram(binwidth = 1) #all overdose, unadjusted
ggplot(data=mcod.st15, aes(x=all.aa)) +
    geom_histogram(binwidth = 1) #all overdose, adjusted

#get summary distribution
summary(mcod.cnty15$all)

#plot histograms (county level outcomes)
ggplot(data=mcod.cnty15, aes(x=all)) +
    geom_histogram(binwidth = .1) #all overdose, unadjusted
ggplot(data=mcod.cnty15, aes(x=all.aa)) +
    geom_histogram(binwidth = .1) #all overdose, adjusted

```

+[1k]Mapping Set up (Using tmap Package)
```{r mapping, eval=TRUE, cache=TRUE}
###--GEOGRAPHIC VARIATION: CHOROPLETH MAPS--###----

#//TODO//#
#compare adjustment
#stratify by drug
#compare by year

#define map function
mkmap <- function(x,x1,y,z){
  #[params --> x==variable to map; x1==option for adjusted; y==year; z==1 for state, 2 for county level]
  if(z==1){
    g<-"State"
    d<-geo.st.mcod
  } else{
    g<-"County"
    d<-geo.cnty.mcod
  }
  if(x1==1){
    a<-".aa"
    b<-"Age-adjusted"
  } else {
    a<-""
    b<-"Unadjusted"
  }
  v<-paste0(x,a,"_",substr(y,3,4))
  tm_shape(d) +
  tm_fill(v, style = "quantile", n = 7, palette = "Greens", 
          title = paste0(b," Rate/1k of Drug-related Deaths(",x,"), by ",g," (",y,")")) + 
  tm_legend(bg.color = "white", bg.alpha = 0.6) + 
  tm_style_gray() #state map, unadjusted, all drugs
}

mkmap("all",1,"2015",1) #state, 2015, all, age adjusted
#mkmap("all",0,"2015",1) #state, 2015, all,unadjusted
#dev.print(png, file = "EX7-drug-all-by-state-2015.png", width = 500, height = 300) #print to file

#mkmap("rx",1,"2015",1) #state, 2015, rx, age adjusted
#mkmap("rx",0,"2015",1) #state, 2015, rx, unadjusted
#mkmap("opioid",1,"2015",1) #state, 2015, opioid, age adjusted
#mkmap("opioid",0,"2015",1) #state, 2015, opioid, unadjusted
#mkmap("heroin",1,"2015",1) #state, 2015, heroin, age adjusted
#mkmap("heroin",0,"2015",1) #state, 2015, heroin, unadjusted

mkmap("all",1,"2015",2) #county, 2015, all, adjusted
#mkmap("all",0,"2015",2) #county, 2015, all, unadjusted
#dev.print(png, file = "EX8-drug-all-by-county-2015.png", width = 500, height = 300) #print to file

#mkmap("opioid",0,"2015",2) #county, 2015, opioid, unadjusted
#mkmap("heroin",0,"2015",2) #county, 2015, heroin, unadjusted

#compare over time (for future iterations with multiple year format)
#mkmap("opioid",0,"2015",2) #county, 2015, opioid, unadjusted
#mkmap("opioid",0,"2010",2) #county, 2010, opioid, unadjusted
#mkmap("opioid",0,"2005",2) #county, 2005, opioid, unadjusted
#mkmap("opioid",0,"2000",2) #county, 2000, opioid, unadjusted
#mkmap("opioid",0,"1995",2) #county, 1995, opioid, unadjusted

```

##PART 2: DISCOVERY
+[2a]Set-up Spatial Analysis (Moran's I)
```{r spatial analysis, eval=TRUE}
###--GEOGRAPHIC VARIATION: TESTING SPATIAL AUTOCORRELATION--###----

#RESOURCE: [Moran's I](http://rspatial.org/analysis/rst/3-spauto.html#compute-moran-s-i)

#define global spatial autocorrelation test----
globalSA <- function(x,x1,x2,x3){
  #[params --> x==variable to map; x1==option for adjusted; x2==year; x3==1 for state, 2 for county level]
  if(x3==1){
    g<-"State"
    d<-geo.st.mcod
  } else{
    g<-"County"
    d<-geo.cnty.mcod
  }
  if(x1==1){
    a<-".aa"
    b<-"Age-adjusted"
  } else {
    a<-""
    b<-"Unadjusted"
  }
  v<-paste0("d$",x,a,"_",substr(x2,3,4))
  
  #set up weight matrix
  w <- poly2nb(d, row.names=d$NAME)
  wm <- nb2mat(w, style='B')
  
  #Get ‘y'
  assign("y",eval(parse(text=v)))
  
  #row standardisation of the weights matrix:
  rwm <- mat2listw(wm, style='W')
  
  # Checking if rows add up to 1
  mat <- listw2mat(rwm)
  apply(mat, 1, sum)[1:15]
  moran.plot(y, rwm)
  
  #compute Moran’s I and do a significance test
  ww <-  nb2listw(w, style='B')

  #test for significance. 
  moran.test(y, ww, randomisation=FALSE)
  
  #Monte Carlo simulation
  moran.mc(y, ww, nsim=99)

}

#compare by drug type, state level, 2015----
globalSA("all",1,"2015",1)
#globalSA("rx",1,"2015",1)
#globalSA("opioid",1,"2015",1)
#globalSA("heroin",1,"2015",1)

#compare adjusted/unadjusted----
#globalSA("all",1,"2015",1)
#globalSA("all",0,"2015",1)

#run for county level----
globalSA("all",0,"2015",2)

#compare over time (for future iterations with multiple year format)----
#globalSA("all",1,"2010",1)
#globalSA("all",1,"2005",1)
#globalSA("all",1,"2000",1)
#globalSA("all",1,"1995",1)


```

+[2b]Spatial Analysis (Getis Gi)
```{r spatial analysis, eval=TRUE}
###--GEOGRAPHIC VARIATION: TESTING SPATIAL AUTOCORRELATION--###----

#RESOURCE: [Local Statistics](http://rspatial.org/rosu/rst/Chapter8.html?highlight=local)

#//TODO//#----
#Compare alternative tests

```

+[2c]Set-up Spatial Analysis (LISA)
```{r spatial analysis, eval=TRUE}
###--GEOGRAPHIC VARIATION: TESTING SPATIAL AUTOCORRELATION--###----

#RESOURCE: [Tutorial](http://rstudio-pubs-static.s3.amazonaws.com/4938_b5fc230d586c48b291627ff6ea484d2e.html)

#define local spatial autocorrelation test function----
localSA <- function(x,x1,x2,x3){
  #[params --> x==variable to map; x1==option for adjusted; x2==year; x3==1 for state, 2 for county level]
  if(x3==1){
    g<-"State"
    d<-geo.st.mcod
  } else{
    g<-"County"
    d<-geo.cnty.mcod
  }
  if(x1==1){
    a<-".aa"
    b<-"Age-adjusted"
  } else {
    a<-""
    b<-"Unadjusted"
  }
  v<-paste0("d$",x,a,"_",substr(x2,3,4))
 
  print(paste(x2,x,b,g,sep=" "))

  #set up weight matrix
  w <- poly2nb(d, row.names=d$NAME)
  wm <- nb2mat(w, style='B')
  
  #Get ‘y'
  assign("y",eval(parse(text=v)))
  
  # create row-standardized Queens contiguity weights matrix
  dt_nbq <- poly2nb(d)  #queen's neighborhood
  dt_nbq_w <- nb2listw(dt_nbq)
  locm <- localmoran(y, dt_nbq_w)  #calculate the local moran's I
  print(summary(locm))
  
  # manually make a moran plot standarize variables
  d$s <- scale(y)  #save to a new column
  
  # create a lagged variable
  d$lag_s <- lag.listw(dt_nbq_w, d$s)
  
  print(summary(d$s))
  
  print(summary(d$lag_s))
  
  plot(x = d$s, y = dt$lag_s, main = " Moran Scatterplot ")
  abline(h = 0, v = 0)
  abline(lm(d$lag_s ~ d$s), lty = 3, lwd = 4, col = "red")
  
  # check out the outliers click on one or twon and then hit escape (or
  # click finish)
  identify(d$s, d$lag_s, d$NAME, cex = 0.8)
  
  # identify the moran plot quadrant for each observation
  d$quad_sig <- NA
  d@data[(d$s >= 0 & d$lag_s >= 0) & (locm[, 5] <= 0.05), "quad_sig"] <- 1
  d@data[(d$s <= 0 & d$lag_s <= 0) & (locm[, 5] <= 0.05), "quad_sig"] <- 2
  d@data[(d$s >= 0 & d$lag_s <= 0) & (locm[, 5] <= 0.05), "quad_sig"] <- 3
  d@data[(d$s >= 0 & d$lag_s <= 0) & (locm[, 5] <= 0.05), "quad_sig"] <- 4
  d@data[(locm[, 5] > 0.05), "quad_sig"] <- 5  
  #ASSIGN A 5 TO ALL NON-SIGNIFICANT OBSERVATIONS
  
  # Set the breaks for the thematic map classes
  breaks <- seq(1, 5, 1)
  
  # Set the corresponding labels for the thematic map classes
  labels <- c("High-High", "Low-Low", "High-Low", "Low-High", "Not Signif.")
  
  # see ?findInterval - This is necessary for making a map
  np <- findInterval(d$quad_sig, breaks)
  
  # Assign colors to each map class
  colors <- c("red", "blue", "lightpink", "skyblue2", "white")
  plot(d, col = colors[np])
  mtext("Local Moran's I", cex = 1.5, side = 3, line = 1)
  legend("bottomleft", legend = labels, fill = colors, bty = "n")
  
  var<-paste0(x,a,"_",substr(x2,3,4))
  assign(paste0(x2,x,a,g),d@data[,c("geo",var,"s","lag_s","quad_sig")], envir = .GlobalEnv)
  
}

#compare by drug type, state level, 2015----
localSA("all",1,"2015",1)
#localSA("rx",1,"2015",1)
#localSA("opioid",1,"2015",1)
#localSA("heroin",1,"2015",1)

#compare adjusted/unadjusted----
#localSA("all",1,"2015",1)
#localSA("all",0,"2015",1)

#run for county level----
#localSA("all",0,"2015",2)
#localSA("rx",0,"2015",2)
#localSA("opioid",0,"2015",2)
#localSA("heroin",0,"2015",2)

#compare over time (for future iterations with multiple year format)----
#localSA("all",1,"2010",1)
#localSA("all",1,"2005",1)
#localSA("all",1,"2000",1)
#localSA("all",1,"1995",1)

```

+[2d]Cluster Identification
```{r unsupervisedmachine learning , eval=TRUE}
###--CLUSTER IDENTIFICATION (UNSUPERVISED METHODS): K-means--###----

#//TODO//#----
#Test a k-means algorithm of outcomes, by county and year to look for clusters in time/space


```

##PART 3: ESTIMATION

+[3a]Import data
```{r load data, eval=TRUE}
###--IMPORT DATA--###----

#building permits----
#source("data/building_permits/building_permits.R")
buildingpermits<-readRDS(file="data/building_permits/buildingpermits.RDa")

#crime----
#source("data/crime/crime.R")
crime<-readRDS(file="data/crime/crime.Rda")

#demographics----
#source("data/demographics/demographics")
demographics<-readRDS(file="data/demographics/demographics.Rda")

#built environment----
#source("data/built_environment/built_environment.R")
ruralurbancodes2013<-readRDS(file="data/built_environment/ruralurbancodes2013.RDa")

#economic----
#to pull from US Census Bureau (https://www.census.gov/economic-indicators)

#election results----
#source("data/election_results/election_results.R")
election_results<-readRDS(file="data/election_results/election_results.Rda")
                        
#gdp----
#source("data/gdp/gdp.R")
per_captia_gdp_by_state_multi_year<-readRDS(file="data/gdp/per_captia_gdp_by_state_multi_year.Rda")
per_industry_contribution_to_state_gdp_multi_year<-readRDS(file="data/gdp/per_industry_contribution_to_state_gdp_multi_year.Rda")

#health behavior----
#to be cleaned -- see ("data/health_behavior")

#healthcare----
#source("data/healthcare/healthcare.R")
healthcare<-readRDS(file="data/healthcare/healthcare.RDa")

#homelessness----
#to be cleaned -- see ("data/homeless/homeless.R")

#hospital compare----
#to be cleaned -- see ("data/hospital_compare")

#jobs----
#to be cleaned -- see ("data/jobs/jobs.R")

#marijuana----
#to be cleaned -- see ("data/marijuana")

#mobility inequality----
#to be cleaned -- see ("data/mobility_inequality")

#mortality----
#source("data/mortality/mortality.R")
mortality<-readRDS("data/mortality/mortality.RDa")

#needle exchange----
#to be cleaned -- see ("data/needle_exchange")

#opioid prescription rates----
#source("data/opioid_prescription_rates/rx_rates.R")
rx_rates<-readRDS("data/opioid_prescription_rates/rx_rates.RDa")
  
#public health----
#to be cleaned -- see ("data/needle_exchange")

#business analyst----
#to be cleaned -- see ("data/business_analyst/business_analyst.R")

```

+[3b]Build Analytic Dataset
```{r merge data, eval=TRUE}
###--MERGE DATA--###----
#source("data/hotspots/hotspots.R")
hotspots.vars <- readRDS(file="data/hotspots/hotspots.suppressed.Rda")

#specify variables of interest
names(hotspots.vars)

#geography
vars.geogr<-c("geo","NAME","STUSPS","state.region")

#outcomes
vars.outco<-c("all.rt","rx.rt","opioid.rt","heroin.rt")
vars.hotsp<-c("all.hh","rx.hh","opioid.hh","heroin.hh","hotspot_any","hotspot_heroin")

#gen pop demographics
vars.demog<-c("female.2012.pct","age.55over.2012","age.mdn.2012","race.wht.2012","ethnicity.2012")

#gen pop education and employment
vars.educa<-c("edu.male.nohs.2012.pct","edu.hsdropout.2012.pct","unemployment.male.2012.pct")

#gen pop socio economic
vars.socio<-c("hhinc.mdn.2012","pov.lt2x.2012.pct","households.singlefemale.2012")

#healthcare
vars.health<-c("healthcare.emdepts.2012.dens","healthcare.trauma.2012.dens","healthcare.adm.2012.rt","healthcare.edvisits.2012.rt","healthcare.hosp.highvol.2012.pct","healthcare.mds.em.2012.rt","healthcare.mds.rsrch.2012.rt")

#health policy
vars.hplcy<-c("insurance.2012.pct","medicaid_expansion")

#public health
vars.pubhlth<-c("homcide.firearm.2012.rt","suicide.2012.rt" )

#crime
vars.crime<-c("crime.2012.rt","murder.2012.rt","robbery.2012.rt")

#built envt
vars.benvt<-c("popdens.2012","vacancy.2012.pct","urbanrural","permits.5yrchng")

#politics
vars.politics<-c("flip0816","dem2rep16")

#economy
vars.econom<-c("foreign.since2000.2012.pctchng","gdp.10yr.growth","gdp15.mxind","gdp10y.chgind")

```

+[3c]Descriptive Analysis
```{r plot data, eval=TRUE}
###--PLOT DATA--###----

#cat vars > bar charts
d<-hotspots.vars
ggplot(data=d, aes(x=ifelse(hotspot_any>0,1,0))) + 
    geom_bar()

ggplot(data=d[hotspot_any>0,], aes(x=hotspot_heroin)) + 
    geom_bar()

ggplot(data=d, aes(x=urbanrural)) +
    geom_bar()

#cont vars > histograms
ggplot(data=d[all.rt>0,], aes(x=all.rt)) +
    geom_histogram(binwidth = .1)


ggplot(d[all.rt>0,], aes(all.rt, fill = state.region)) + geom_density(alpha = 0.2)
ggplot(d[rx.rt>0,], aes(rx.rt, fill = state.region)) + geom_density(alpha = 0.2)
ggplot(d[heroin.rt>0,], aes(heroin.rt, fill = state.region)) + geom_density(alpha = 0.2)
ggplot(d[opioid.rt>0,], aes(opioid.rt, fill = state.region)) + geom_density(alpha = 0.2)

ggplot(d[hotspot_heroin>0,], aes(age.mdn.2012, fill = factor(hotspot_heroin))) + geom_density(alpha = 0.2)

ggplot(d[rx.rt>0,], aes(rx.rt, fill = state.region)) + geom_density(alpha = 0.2)
ggplot(d[heroin.rt>0,], aes(heroin.rt, fill = state.region)) + geom_density(alpha = 0.2)
ggplot(d, aes(all.rt, fill = medicaid_expansion)) + geom_density(alpha = 0.2)

ggplot(d, aes(hhinc.mdn.2012, fill = as.factor(hotspot_heroin))) + geom_density(alpha = 0.2)
ggplot(d, aes(gdp.10yr.growth, fill = as.factor(gdp15.mxind))) + geom_density(alpha = 0.2)

#bin/cat x bin/cat vars >> bar charts

#Proportions of each "fill" category within each x-axis category:
ggplot(data=d, aes(x=hotspot_heroin, fill=urbanrural)) +
    geom_bar(position="fill")

#Counts for each "fill" category side-by-side for each x-axis category:
ggplot(data=d, aes(x=hotspot_any, fill=state.region)) +
    geom_bar(position="dodge")

#cont var x bin/cat >> box plots
ggplot(data=d, aes(x=factor(state.region), y=all.rt)) +
    geom_boxplot()

#Two cont vars >> scatter plots
ggplot(data=d, aes(x=popdens.2012, y=heroin.rt, colour=factor(hotspot_heroin))) +
    geom_point()

```

+[3a]Regression Analysis
```{r linear models, eval=TRUE}
###--LINEAR REGRESSION--###----

#subset only counties with at least 1 drug death
d <- d[all.rt>0,]

#define batch plotting functions
batchplot <- function(x1,y1){
  lapply(x1,function(a){
  assign("x",eval(parse(text=paste0("d$",a))))
  assign("y",eval(parse(text=paste0("d$",y1))))
  ggplot(d, aes(x=x, y=y)) + xlab(a) + ylab(y1) +
    geom_point(color="blue")
  }
)
}

batchbox <- function(x1,y1){
  lapply(x1,function(a){
  assign("x",eval(parse(text=paste0("d$",a))))
  assign("y",eval(parse(text=paste0("d$",y1))))
  ggplot(d, aes(x=x, y=y)) + xlab(a) + ylab(y1) +
    geom_boxplot(color="blue")
  }
)
}

#plot death rates x demographics
batchplot(vars.demog,"all.rt")

#plot death rates x education/employment
batchplot(vars.educa,"all.rt")

#plot death rates x socioeocnomic
batchplot(vars.socio,"all.rt")

#plot death rates x healthcare
batchplot(vars.health,"all.rt")

#plot death rates x insurance 
batchplot(vars.hplcy[1],"all.rt")

#plot death rates x medicaid expansion
batchbox(vars.hplcy[2],"all.rt")

#plot death rates x homicide/suicide
batchplot(vars.pubhlth,"all.rt")

#crime
batchplot(vars.crime,"all.rt")

#built envt
batchplot(vars.benvt[c(1,2,4)],"all.rt")
batchbox(vars.benvt[3],"all.rt")

#politics
batchplot(vars.politics[2],"all.rt")
batchbox(vars.politics[1],"all.rt")

#economy
vars.econom<-c("foreign.since2000.2012.pctchng","gdp.10yr.growth","gdp15.mxind","gdp10y.chgind")
batchplot(vars.econom[c(1,2)],"all.rt")
batchbox(vars.econom[c(3,4)],"all.rt")

#make formula of variable names by group vector
mkformula<-function(x){
  y<-x[1]
  for(i in x){
    y<-paste(y,i,sep=" + ")
  }
  return(y)
  }



summary.lm(lm(all.rt ~ race.wht.2012 + female.2012.pct, data=d))
summary.lm(lm(all.rt ~ race.wht.2012 + female.2012.pct + race.wht.2012:female.2012.pct, data=d))

summary(lm(eval(parse(text=paste0("all.rt~",mkformula(vars.demog)))), data=d))
summary(lm(eval(parse(text=paste0("all.rt~",mkformula(vars.socio)))), data=d))
summary(lm(eval(parse(text=paste0("all.rt~",mkformula(vars.educa)))), data=d))
summary(lm(eval(parse(text=paste0("all.rt~",mkformula(vars.health)))), data=d))
summary(lm(eval(parse(text=paste0("all.rt~","factor(urbanrural)"))), data=d))


```

+[3b]Regression Analysis
```{r logistic, eval=TRUE}
###--LOGISTIC REGRESSION--###----

d$hotspot_bin <- ifelse(d$hotspot_any>0,1,0)

chisq.test(table(d$hotspot_bin, d$urbanrural)) 
summary((glm(hotspot_bin~urbanrural, data=d, family=binomial())))
exp(coef(glm(hotspot_bin~urbanrural, data=d, family=binomial()))) #Odds ratios


```

##PART 4: PREDICTION

+[4a]Regression Analysis
```{r surpervised learning, eval=TRUE}
###--RANDOM FORREST--###----

dsub <- data.table(sapply(d[,21:104],function(x)ifelse(is.na(x),0,x)))

dt.rf <- randomForest(hotspot_bin~., data=dsub, ntree=100, importance=TRUE) #run random forest function
    summary(dt.rf$importance[,4]) #print summary of importance measures
    quantile(dt.rf$importance[,4], prob = seq(0, 1, length = 11), type = 5) # list deciles of MeanDecreaseGini values
    rf.decile <- quantile(dt.rf$importance[,4], c(.90))
    dt.rf.important <- as.vector(names(dt.rf$importance[dt.rf$importance[,4]>rf.decile,c(0,4)])) # store var names in top decile as vector
    'The following variables were identified as most important, based on the top decile of MeanDecreaseGini values:'
    dt.rf.important #print variables with importance in top decile
    dt.rf #report summary to get OOB
    dt.rf.pred <- predict(dt.rf, dsub, type="prob") #predict outcome/classification
    dt.rf.pred.control <- dt.rf.pred[, 2] #extract predicted probability of 'control' only


```

##PART 5: VISUALIZATION

+[5a]Visualization
```{r plot data, eval=TRUE}
###--HEAT MAPS--###----

#get gini values, input into matrix
dmat<-t(data.matrix(d[,21:103]))
heatmap <- heatmap(dmat, Rowv=NA, Colv=NA, col = heat.colors(256), scale="column", margins=c(5,10))

my_palette <- colorRampPalette(c('red','yellow','green'))(256)

scaled <- scale(dmat[,-1])    # scale all but the first column to make information comparable
heatmap.2(scaled,               # specify the (scaled) data to be used in the heatmap
          cexRow=0.5, 
          cexCol=0.95,          # decrease font size of row/column labels
          col = my_palette,     # arguments to read in custom colors
          colsep=c(2,4,5),      # Adding on the separators that will clarify plot even more
          rowsep = c(6,14,18,25,30,36,42,47), 
          sepcolor="black", 
          sepwidth=c(0.01,0.01),  
          scale="none",         # we have already scaled the data 
          dendrogram="none",    # no need to see dendrograms in this one 
          trace="none") 

```